The effects of uncertain context inference on motor adaptation

### 2022-03-31 adaptation/article/article.tex
Finished COIN-based model description

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}%article}
% \documentclass{article}
\usepackage[font=small]{caption}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{todonotes}
\usepackage{authblk}
% \usepackage[nomarkers,figuresonly]{endfloat}

@@

\section{Methods}
\subsection{Model description}
\label{subsection:model-description}
We first describe the general principle behind the model, followed, in the next
subsection, by the full mathematical description of each one of its components.

Before proceeding with the model description, let us introduce nomenclature:
\begin{itemize}
\item $t$: Discrete time at the moment an observation is made, before inference is done. 
Here taken as the trial number.
\item $\vartheta_t$: Observation. Note that this is assumed to be a noisy
observation of the generalized state (see below)
\item $s_t$: Hidden state, e.g. the position of the hand in reaching
experiments. It does not depend on the context.
\item $c_t$: Motor command (action) taken after observing $\vartheta_t$ and inferring the context.
\item $\zeta_t$: Categorical variable representing the context inferred after observing $\vartheta_t$.
\item $\omega_t$: Contextual information. This term includes, e.g. visual cues,
proprioceptive feedback, performance feedback.
\item $z_t$: Generalized state, $z_t = \{s_t, \zeta_t\}$. Observed noisily
as $\vartheta_t$.
\item $m_i$: Forward model for context $\zeta_i$.
\item $\phi_i$: Parametrization of the forward model $m_i$. In adaptation experiments, this
can be interpreted as the inferred size of the adaptation.
\end{itemize}

Context inference can work through context-specific (possibly unreliable) sensory cues,
or through the dynamics of the system (e.g. expected outputs of motor
commands), using prediction error as the force behind inference. In this case,
the generative model can be written as follows:
\[
p(z_t, \vartheta_t | z_{t-1}, c_{t-1}) = p(\vartheta_t | z_t)p(z_t|z_{t-1}, c_{t-1})
\]
and inferring the current (hidden) generalized state is done with Bayes' theorem:
\[
  q(z_t | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | z_t)p(z_t|\vartheta_{t-1}, c_{t-1})
\]

Additionally, after having inferred the current state, motor adaptation is given by
\begin{equation}
q(\phi | z_t, z_{t-1}, c_{t-1}) \propto p(z_t | z_{t-1}, c_{t-1}, \phi)q(z_{t-1}|\ldots)p(\phi) \label{eqn:motor-adaptation}
\end{equation}
where $\phi$ are the parameters of the internal model for the dynamics, also called the forward model,
i.e. the function $f: (z_t, c_t) \rightarrow z_{t+1}$, which determines the beliefs of
the agent regarding how the system evolves after having taken action $c_t$. $q(z_{t-1}|\ldots)$ is the
posterior probability calculated in the previous trial.

Having inferred the generalized state (including context), a motor command is
sampled from a distribution:
\begin{equation}
p(c_t | s_t, \zeta_t) = \displaystyle\sum_{i}q(\zeta_i)p(c_t | s_t, \zeta_i)
\end{equation}
where $i$ runs over all contexts and $p(a_t | ... m)$ is the probability distribution over actions
prescribed by the forward model $m_i$ for context $\zeta_i$.

In the following sections, we present a specific mathematical implementation
of the principles outlined above. This is the implementation used throughout the
simulations in the Results section.

\subsubsection{Mathematical implementation}
The generalized state is given by:
\begin{equation}
z_t = \{s_t, \zeta_t\}
\end{equation}
where $s_t$ is, e.g. in a reaching task, the current position of the hand in
Cartesian coordinates, where the origin is the starting point. We assume that
motor commands are issued every $\Delta t$ for simplicity. At the beginning of
each time interval, the context is inferred combining the contextual
information and the prediction error of the outcome of the previous motor
command. This yields:
\begin{align}
  \begin{split}
  q(\zeta_t) &= q(\zeta_t | \omega_t)q(\zeta_t | s_t, s_{t-1}, c_{t-1}) \\ \label{eqn:estimated-context}
  q(\zeta_t | \omega_t) &\propto p(\omega_t | \zeta_t)p(\zeta_t) \\
  q(\zeta_t | s_t, s_{t-1}, c_{t-1}) &\propto p(s_t | \zeta_t, s_{t-1}, c_{t-1})p(\zeta_t)
  \end{split}
\end{align}
The terms $p(\zeta_t)$ refer to the prior probability of the context at the
beginning of the trial: at the first trial, this refers to prior beliefs over
which contexts are more common/likely, which we call $p_0(\zeta_t)$. At each
subsequent trial, it incorporates the belief that has so far been accumulated,
given previous trials. Note that certain events, such as the start of a new
block of trials (see main text) can return an agent to a state in which
$p_0(\zeta_t)$ becomes relevant.

We do not directly model hand movements or eye position. Instead, we model
systems in the motor error space: the position $s_t = 0$ represents a trial in
which no motor error occurred. We make use of the fact that in most
experiments, motor adaptation needs to happen in one of two directions
(e.g. clockwise vs. counter-clockwise) to keep the error space
one-dimensional. With this, $s_t > 0$ represents motor error in one direction
(e.g. clockwise) and $s_t < 0$, error in the other direction. We made this
choice because in this work we modeled different experiments with
different state spaces, but with the commonality of (at most)
two-dimensional error spaces. In order to model a specific experiment in its
own physical space (e.g. hand positions in a reaching task), only the dynamical
system  $p(s_t | \zeta_t, s_{t-1}, c_{t-1})$ would need to be adapted.

We assume that the decision-making agent has a baseline internal forward model $m_O$ that
maps motor commands to outcomes:
\begin{equation}
p(s_{t+1} | s_t, c_t, m=m_O) = f_O(s_t, c_t)
\end{equation}
where $c_t$ are motor commands (actions). We further assume that participants
have learned $f_O$ throughout their lives and use it to decide which motor
command to issue under normal circumstances (i.e. no adaptation). When a
perturbation is introduced, to which the participant must adapt, we assume they
learn a new forward model $f_m^{\phi}$, where $\phi$ represents the parameters
of the function that participants must learn. For example, in a saccade task,
$\phi$ represents the new gain. Note that $\phi$ depends on the context and
participants may need to use and update multiple forward adapted models
$f_m^{\phi}$ throughout one experiment.

Given a forward model $f_m$ (or $f_m^\phi$), action selection is done by
building a probability distribution over possible actions $c_t$ based on how
likely they are to produce zero motor error. For example, in saccade
experiments, motor commands are issued based on how likely they are to move the
fovea to the target. These motor commands are issued to counteract intrinsic
sources of error, such as muscle variability and exhaustion, and extrinsic
sources of error, such as the experimental manipulations on saccade gain. Thus,
the expected observed error can be written as:
\begin{equation}
p(\epsilon | c_t) \propto p(\epsilon_{\text{intrinsic}} | c_t) + p(\epsilon_{\text{extrinsic}} | c_t)
\end{equation}
and a motor command can be chosen to maximize the chance of observing zero error:
\begin{equation}
p(c_t) = argmin_{c_t} p(\epsilon | c_t)
\end{equation}

For simplicity, we assume that motor errors are Gaussian, such that:
\begin{equation}
p(\epsilon | c_t) = N(\mu_{\text{extrinsic}}, \sigma_\epsilon)
\end{equation}
where $\sigma_\epsilon$ incorporates both intrinsic and extrinsic sources of
error and is a free parameter of the model. The mean of the error is assumed to
have no intrinsic component.

With this, a forward model will produce a motor command with outcomes centered
at $-\mu_{\text{extrinsic}}$, which counteracts the externally-induced errors.

To complete the action-selection picture, we return to context inference. As
discussed above, action selection is made via a weighted sum of the motor commands
produced by each forward model:
\begin{equation}
p(c_t | ...) \propto \displaystyle \sum_{i} q(\zeta_i)p(c_t | \zeta_i, ...) \label{eqn:dist-comm}
\end{equation}
From this distribution, an action is sampled at each trial.

To update the agent's estimates on the parameters of the forward models, we
make use of Bayes' theorem with conjugate priors. This approach has the
advantage of not only greatly simplifying computational calculations, but also
presenting update equations with intuitive appeal.

In the following equations, we drop the dependence on the context and
observations for notational simplicity:
\begin{equation}
q(\phi | s_{1:t}) \propto p(s_{1:t} | \phi)p(\phi)
\end{equation}
where $s_{1:t}$ are the observed states up until time $t$. The likelihood is a
Gaussian:
\begin{align}
  p(s_t | \phi) &= N(\mu_s, \sigma_s)  \\
  \mu_s &= \phi + s_t \\
  \sigma_s &= \xi_0 1
\end{align}
This likelihood is the probability of observing $s$ given the previous estimate
of the parameters of the forward model, $\phi_{t-1}$, and given the dynamics of
the system (see below).

To update $\phi$, we assumed that for each context, the agent's belief over
the magnitude of the adaptation is given by a normal distribution:
\begin{equation}
p(\phi) = N(\mu_\phi, \sigma_\phi) \label{eqn:data-dist}
\end{equation}
where $\mu_\phi$ and $\sigma_\phi$ are parameters to be estimated at each
trial. Before the experiment begins, the agent will have a prior distribution
over these parameters. A standard Bayesian approach is to choose NormalGamma
priors:
\begin{equation}
p(\mu_\phi, \sigma_\phi) = NG(\mu_0, \nu_0, \alpha_0, \beta_0)
\end{equation}
where $\mu_{\phi}^0, \nu_{\phi}^0, \alpha_{\phi}^0$ and $\beta_{\phi}^0$ are free
hyperparameters of the model.

The update equations for the magnitude parameters are given by:
\begin{equation}
q(\mu_\phi, \sigma_\phi | s_t, s_{t-1}, c_{t-1}) \propto p(s_t |
s_{t-1}, c_{t-1}, \zeta_t)p(\mu_\phi, \sigma_\phi) \label{eqn:context-from-x}
\end{equation}
where $p(s_t | s_{t-1}, c_{t-1}, \zeta_t)$ is a Gaussian distribution centered
around $\mu_c$ with standard deviation $\sigma_c$, which is a free parameter of
the model. It is assumed that the standard deviation over outcomes ($\sigma_c$)
is related to that of the parameter $\mu_\phi$ as follows:
\begin{equation}
\sigma_\phi = \sigma_c / \nu_\phi
\end{equation}
which makes $\nu_\phi$ a parameter to estimate. Each one of the forward models can
be updated at every trial, depending on context inference (see main text). The
updated parameters for any one forward model $i$ are given by:
\begin{align}
  \begin{split}
  \mu_\phi^{(t)} &= \frac{\nu_\phi^{(t-1)} \mu_\phi^{(t-1)} + q(\zeta_i)s_t}{\nu_\phi^{(t-1)} + q(\zeta_i)} \\
  \nu_\phi^{(t)} &= \nu_\phi^{(t-1)} + q(\zeta_i) \\
  \alpha_\phi^{(t)} &= \alpha_\phi^{(t-1)} + q(\zeta_i) / 2 \\
  \beta_\phi^{(t)} &= \beta_\phi^{(t-1)} + \frac{q(\zeta_i)\nu_\phi^{(t-1)}}{\nu_\phi^{(t-1)} +
    q(\zeta_i)}\frac{\left(s_t - \mu_\phi^{(t-1)}\right)^2}{2}  \label{eqn:update-full}
  \end{split}
\end{align}
where $q(\zeta_i)$ is the posterior probability of context $\zeta_i$ from
\eref{eqn:estimated-context}. Note that when $q(\zeta_i) = 0$, the model $i$ is not
updated.

>>>>

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}%article}
% \documentclass{article}
\usepackage[font=small]{caption}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{todonotes}
\usepackage{authblk}
% \usepackage[nomarkers,figuresonly]{endfloat}
\usepackage{multirow}
\usepackage{tabularx}

@@

\section{Methods}
\subsection{The COIN model}
In this work, we used the recently-introduced COIN model \citep{Heald_Contextual_2021}, adapted to the experiments that we covered in our simulations. In this section, we give a brief introduction to the COIN model and, in the subsequent subsection, describe how we adapted the model to the experimental tasks. For a full description of the model, refer to \cite{Heald_Contextual_2021}.

\subsubsection{Generative model}
At each trial $t$, the agent infers both the context and the context-dependent adaptation (e.g. the parameters of the force field in mechanical-arm experiments). The context is represented by a latent, categorical variable $\zeta_t$:
\begin{equation}
p(\zeta_t | \zeta_{t-1}, \pi_{\zeta_{t-1}}) = \text{Discrete}\left(\pi_{\zeta_{t-1}}\right)
\end{equation}
where $\pi_{\zeta_{t-1}}$ is the transition probability vector from context $\zeta_{t-1}$ to all other contexts. MARKOV CHAIN. The contextual cues (when present in an experiment) are assumed to be drawn depending on the context following:
\begin{equation}
p(q_t | c_t, \Phi) = \text{Discrete}(\Phi_{\zeta_t})
\end{equation}
where $Q_{\zeta_t}$ is the probability vector with which the contextual cue $q_t$ is shown to the agent in context $\zeta_t$. As pointed out by \cite{Heald_Contextual_2021}, both $\Phi$ and $\pi$ are in principle infinite, but a task-relevant finite set can be used instead.

The context-dependent adaptation is represented by the latent variable $x_{\zeta,t}$ and assumed to arise from an autoregressive process AR(1), with a stationary Gaussian distribution of unknown mean and variance:
\begin{equation}
p(x_{\zeta,t}) = \mathcal{N}(\mu_{\zeta,x}, \sigma_{\zeta,x})
\end{equation}
where $\sigma$ is assumed to be shared across all contexts. Note that $\mu$ and $\sigma$ are parametrized by the parameters of the AR(1) process, namely $\mu_{\zeta,x} = d_\zeta / (1 - a_\zeta)$ and $\sigma_{\zeta, x} = \sigma_q / 1 - a_\zeta^2$, where $\sigma_q$ is a free parameter of the model.

Observations take the form of state feedback (e.g. the position of the cursor on the screen in visuomotor rotation tasks), given by:
\begin{equation}
y_t = x_{\zeta_t, t} + \nu_t
\end{equation}
where $\nu_t$ is a zero-mean Gaussian noise term with unknown standard deviation $\sigma_r$, which is a free parameter of the model.

Action selection (i.e. motor output $u_t$) is done via the weighted mean of $x_{j,t}$:
\begin{equation}
u_t  = \displaystyle\sum_{j}p(\zeta_{j,t} | ...) x_{j,t}
\end{equation}
To include motor noise (independent from estimation uncertainty), as well as carry over the uncertainty over $x_{j,t}$, we instead sample motor commands from a Gaussian centered on this mean, with a standard deviation $\sigma_u$, which is a free parameter of the model.

\subsubsection{Using the model}
The free parameters of this model can be fit to participants' data, as was done by \cite{Heald_Contextual_2021}. In this work, we instead chose values for these parameters to show that the model is capable of explaining the experimental phenomena in the Results section. Additionally, by fixing these parameters the agent is able to perform exact Bayesian inference at each trial using conjugate priors, replacing the MCMC approach used by \cite{Heald_Contextual_2021} due to the mathematical intractability of the full formulation. This, however, does not change the model and was done purely for computational efficiency. In this section, we describe how we fixed parameters and the procedure for Bayesian inference.

As explained above, context is assumed to be a discrete variable which evolves as a Markov process. The transition matrices $\pi$ were generated via a Dirichlet process, with parameters that can be inferred from participant data ($\alpha$ and $\kappa$ in \cite{Heald_Contextual_2021}). For a fixed value of these parameters, the transition matrices also become fixed. In particular, setting the parameters to CALCULATE THIS, the transition matrices are such that contexts are assumed to be sticky, i.e. a transition from a context to itself is more likely than to any other context. In our simulations, we set the probability of self-transitioning (denoted $p_\zeta$) depending on the experiments (see below), to numbers that match the information given to the participants of each study.

Contextual cues are assumed by the agent to be sampled from a distribution that depends on the current context. This is done through a set of cue probability vectors that are generated via a parametric distribution, whose parameters are fitted to participants' data. Because the experiments we simulate do not include probabilistic or deceiving cues, contextual cues, when present, unequivocally reflect the current context, i.e. $p(q_t = i | c_t = j) = d_{ij}$, where $\delta_{ij}$ is the Kronecker delta, equaling zero unless $i = j$.

Finally, we chose an equivalent parametrization for the priors over the hidden variables $x_{j,t}$, using the mean $\mu_x$ and standard deviation $\sigma_x$ directly instead of the AR(1) $a$ and $d$ parameters used by \cite{Heald_Contextual_2021}. Note that we dropped the $j$ dependency for clarity. This parametrization, in conjunction with the fixed parameters outlined above,  allows us to set priors over $\mu_x$ and $\sigma_x$ that enable exact inference over the latent variables. The priors are given by a Normal-Gamma distribution, typically used for priors of the mean and standard deviation of a Gaussian distribution [CITATION NEEDED]:
\begin{equation}
\mu_x, \sigma_x \sim \mathcal{NG}(\mu_0, \nu_0, \alpha_0, \beta_0)
\end{equation}
with free parameters $\mu_0$, $\nu_0$, $\alpha_0$ and $\beta_0$, which we fixed for each experiment separately. Because $x$ is context-specific, so are these parameters. This formulation comes with four free parameters (i.e. the hyper-priors $\mu_0, \nu_0, \alpha_0, \beta_0$), in accordance with the original formulation (note that \cite{Heald_Contextual_2021} fixed the mean of the priors for $b$ to zero). While the two formulations are not mathematically identical, the effects of the hyper-priors for both are the same; we discuss these effects in the next section. Additionally, we found numerically that these two formulations could provide almost identical samples for given parametrizations (see SUPPLEMENTARY MATERIALS). \todo{Do this}

Because the likelihood function $p(y_t | x_t, ...)$ is Gaussian, this choice of priors allows us to calculate the update equations as follows:
\begin{align}
  \begin{split}
  \mu_\phi^{(t)} &= \frac{\nu_\phi^{(t-1)} \mu_\phi^{(t-1)} + q(\zeta_i)s_t}{\nu_\phi^{(t-1)} + q(\zeta_i)} \\
  \nu_\phi^{(t)} &= \nu_\phi^{(t-1)} + q(\zeta_i) \\
  \alpha_\phi^{(t)} &= \alpha_\phi^{(t-1)} + q(\zeta_i) / 2 \\
  \beta_\phi^{(t)} &= \beta_\phi^{(t-1)} + \frac{q(\zeta_i)\nu_\phi^{(t-1)}}{\nu_\phi^{(t-1)} +
    q(\zeta_i)}\frac{\left(s_t - \mu_\phi^{(t-1)}\right)^2}{2}  \label{eqn:update-full}
  \end{split}
\end{align}
Note that the effect of the evidence (i.e. observations) on the inference over the context-dependent hidden states is gated by the probability of each context $p(\zeta_i)$, as in \cite[][supplementary materials]{Heald_Contextual_2021}.

\subsubsection{Model parameters}
\tref{tab:parameters} lists all the parameter values that we used during our simulations. The parameters are divided into two categories: (1) task parameters, which encode the way we simulated the experimental design; (2) agent parameters, which correspond to the free parameters listed in the previous section. The variable names for the model parameters are given in the ``Var'' column, corresponding to the variables in the previous section. The values are divided into experiments and, within experiments, into the different groups or conditions that we simulated. 
\begin{table}[h!]
\resizebox{\textwidth}{!}{%
  \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|}
  \cline{1-11}
  & & & Kim (2015) & \multicolumn{2}{c|}{Oh (2009)} & \multicolumn{2}{c|}{Davidson (2004)} & \multicolumn{3}{c|}{Vaswani (2013)} \\
  \cline{2-11}
  & Var & Description & & Exp. 1 & Exp. 2 & Grp. 3A & Grp. -A & Grp. 1 & Grp. 2 & Grp. 3 \\
  \cline{1-11}\cline{1-11}
  \multirow{4}{*}{Task pars.} & & Contextual cues & Yes & \multicolumn{7}{c|}{No} \\
  \cline{2-11}
  & $x_j^*$ & Adaptation sizes & 0, 40, -40 & 0, 20 & 0, 10 & 0, 4, -4 & 0, 4, 12 & 1 & 0, 1 & -0.5, 1 \\
  \cline{2-11}
  & & Adaptation noise & 0.01 & \multicolumn{2}{c|}{1} & \multicolumn{2}{c|}{0.5} & \multicolumn{3}{c|}{0.01} \\
  \cline{2-11}
  & $\sigma_r^*$ & Obs. noise & 3 & \multicolumn{2}{c|}{1.5} & \multicolumn{2}{c|}{0.1} & \multicolumn{3}{c|}{0.01} \\
  \cline{1-11}
  \cline{1-11}
  \multirow{7}{*}{Agent pars.} & $p_\zeta$& Context self-transition & 0.9 & \multicolumn{2}{c|}{0.98} &  \multicolumn{2}{c|}{0.98} & \multicolumn{2}{c|}{0.9} & 0.8 \\
  \cline{2-11}
  & $\mu_0$ & \multirow{4}{*}{Hyper priors} & 0, -1, 1 & \multicolumn{2}{c|}{0, 0} & 0, 4, -4 & 0, 4, 12 & \multicolumn{2}{c|}{0, 1} & 0, 1, -0.5 \\
  \cline{2-2} \cline{4-11}
  & $\nu_0$ & & 1e4, 10, 10 & \multicolumn{2}{c|}{0, 0} & \multicolumn{2}{c|}{1e4, 0.5, 0.5} & \multicolumn{2}{c|}{1e4, 1} & 1e4, 1, 1 \\
  \cline{2-2} \cline{4-11}
  & $\alpha_0$ & & 25e4, 0.25, 0.25 & \multicolumn{2}{c|}{22e4, 2.2} & \multicolumn{2}{c|}{33e4, 33, 33} & \multicolumn{2}{c|}{5e5, 5} & 15e5, 5, 5 \\
  \cline{2-2} \cline{4-11}
  & $\beta_0$ & & 1e5, 2, 2 & \multicolumn{2}{c|}{1e5, 20} & \multicolumn{2}{c|}{1e5, 200, 200} & \multicolumn{2}{c|}{1e5, 2} & 1e5, 2, 2 \\
  \cline{2-11}
  & $\sigma_u$ & Motor noise & 0.001 & \multicolumn{4}{c|}{2} & \multicolumn{3}{c|}{0.17} \\
  \cline{2-11}
  & $\sigma_r$& Obs. noise & 3 & \multicolumn{2}{c|}{3.5} & \multicolumn{2}{c|}{2.5} & \multicolumn{3}{c|}{0.1} \\
  \cline{1-11}
  \end{tabular}}
\caption{Model and simulation parameters.}
\label{tab:parameters}
\end{table}

We estimated the task parameters from the information provided in their respective publications; when direct information was not provided, we estimated it from the reported results; these estimations are not exact, but function as a proof of context. Agent parameter values are held constant for the different conditions or groups for each experiment, except those parameters that are expected to vary across conditions.

The context-dependent hyper-priors $\{\mu_0, \nu_0, \alpha_0, \beta_0\}$ effect a learning rate and, as such, were set such that the learning rate for the baseline context is nearly zero, following the assumption that participants will not change the way they perform these movements outside of the lab. The values for contexts outside of baseline were set identically within each experiment.

The agent has direct knowledge on the adaptation and observation noises of the task; we do not expect this to be the case in an experiment (instead participants would estimate it during their early trials). \todo{check simulations to see if I can make them identical; if not, change this}

The star notation (e.g. $x_j^*$) denotes the real value used in the simulation of the task, which may be different from that assumed by the agent.
